Management induced changes of soil organic carbon on global croplands

### 2020-12-11 SOCbudgetPaper/discussion.Rmd
small add on

## Including agricultural management data changes the sign of the trend

This study provides an analysis of historic SOC stock changes on cropland. We determine the SOC trends on cropland compared to a counterfactual scenario with a world under natural vegetation under identical historical climatic conditions ($SOC_{natveg}$). Our results show, that human activity lead to cumulative SOC emission of around 37 GtC in 2010. Whereas recent modeling estimates of global SOC emissions indicate an ongoing increase of SOC emissions ([@pugh_simulated_2015], [@sanderman_soil_2017]), our study indicates that the global SOC gap is slowly closing due to improved management.

According to [@sanderman_soil_2017] cumulative SOC emissions since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. Our estimate of 37 GtC in 2010 for cropland emissions only is twice as high as [@sanderman_soil_2017] estimations. However, there are large uncertainties in modeling SOC at the global scale, and the [@sanderman_soil_2017] results have been pointed out as potentially conservatively low compared with experimental results, leaving our results in an aceptable range.

Furthermore, [@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to different management systems at all. [@pugh_simulated_2015] considered management effects like tillage and residue recycling in a static way, but neither changes over time nor alignment to observed historical data like yields-levels or no-tillage areas were taken into account. The study moreover concludes that crop productivity gains (increasing yield levels by 18%) do not lead to a substantial decrease in SOC emissions, accounting for only less than 1% change in SOC emissions.

Our study for the first time uses a dynamic management dataset as driver for SOC dynamics. We show that the moderate global cropland expansion of around 11% between 1974 and 2010 and the resulting depletion of SOC stocks in converted cropland has been out weighted by improved agricultural yields and practices. This challenges [@pugh_simulated_2015] findings of only small effects due to improved practices.

Moreover, our sensitivity analysis indicates that (1) recycled residue biomass increases driven by higher yields as well as higher recycling rates play the most important role, followed by (2) improved manure recycling (e.g. due to improved animal waste management systems) and (3) the adoption of no tillage practices.

modeling management effects on the global scale comes however with parametric and structural uncertainties. As pointed out by [@keel_large_2017] and [@smith_how_2020], carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. [@keel_large_2017] question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries which might not be reflected in outdated data sources. Following this line of argumentation, SOC results calculated in this study may overestimate actual SOC stocks due to an overestimation in residue biomass, for which the ratio compared to the harvested organ is normally reduced due to breeding. However, with an evaluation of management effects (see below), there is no indication for a systematic overrating of residue biomass. Rather, it is more likely that we are still missing carbon inputs to the soil.

## Modeled management effect in line with default IPCC assumptions

To validate our modeled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors (@ipcc_2006_2006, @ipcc_2019_2019) which are based on measurement data for croplands (see \@ref(tab:SCFtable)). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones.

```{r captionSCFtable, echo=FALSE}
captionSCFtable <- "This table shows different estimates of stock change factors $F_{SCF}$, defined as the SOC stock of managed croplands relative to the counterfactual SOC stock of undisturbed natural vegetation. We compare average values for four IPCC climate zone classifications. IPCC 2006 and 2019 default factors (medium input) as well as altered factors for low input of organic matter without any other subsystem consideration are compared to results for 1990 and 2010 of this study (SOC budget)."
```

@@

Our estimates correspond very well to the default stock-change factors used for the tier 1 estimation of IPCC, 2006. For the tropical regions the assumptions changed notably from the guidelines in 2006 to the update in 2019, leaving our results too low in comparison with IPCC, 2019. Considering yield gaps in mainly developing regions in the tropics the default assumption of medium input systems might be an overestimation of actual SOC state. The additional effect of considering low inputs of biomass in tropical regions can however not explain the full mismatch to IPCC 2019 values but account for 14-21% of it.

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of $F_{SCF}$.

## SOC stocks inline with literature

The world's SOC stock and its changes are highly uncertain (cite), which is seen in the wide range of global SOC stock estimates (see \@ref(tab:SOCtable)).

```{r captionSOCtable, echo=FALSE}
captionSOCtable <- "modeled as well as data based estimation for global SOC stock in GtC for the first 30 cm of soil aggregated over all land area. Note that SoilGrids, GSOC and WISE do not consider changes over time and rely on soil profile data gather over a long period of time, which makes it hard to pinpoint a specific year for these SOC estimations. In this context they will be compared to modeled data (LPJmL4, this study) for the year 2010."
```

@@

The global estimates of SOC stock from this study are on the lower end compared to other modeled results or more data driven estimates. Looking at regional results \@ref(append:regcompare), our estimates turn out to be in good agreement for most regions, with the largest deviations for boreal areas. Considering that the model was parametrized for croplands, these mismatches are not suprising as the temperature effects on decomposition are fundamentally different for permafrost soils. To avoid that this bias influences our results, our study focuses exclusively on cropland soils, excluding most of the boreal zone. Moreover, when focussing on SOC changes on cropland, pristine natural vegetated areas without human land management under the same historic climatic conditions cancel out in the calculation of SOC emissions.

Our estimates for total SOC stocks of the world, as well as our SOC initialization, are dominated by the representation of natural lands and pastures, which are however only estimated in a basic manner. For example, we do not have a differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass type biomes. This leaves carbon inputs and decay behavior for natural land and pastures rather uncertain. The absolute values of stocks and emissions from land-use change therefore have to be used with caution. Especially in less forested areas the natural land representation higher uncertainty may arise from the parametrization assumptions of natural litterfall.

Nevertheless, total SOC stocks are in a reasonable range and results for SOC stock changes in relative or absolute numbers are not altered by pristine vegetated areas at all. The SOC gap from changed management is only affected by the natural vegetation representation in the case of land use transitions. In this case, former natural land SOC stocks decline. However, as pointed out before, due to low cropland expansion rates, land use change-related SOC declines are outweighed by the general trend of increasing SOC stocks on existing cropland due to management improvements.
We conducted a sensitivity analysis for a wide range of possible parameter combinations for lignin and nitrogen parametrization of natural litterfall, which tend to change the global SOC stock substantially for different choices
(see appendix). This shows that the general trend of decreasing SOC gap is not altered even under very high estimates for natural SOC stocks.

## Important shortcomings --> might be added as well

The **initialization of SOC stocks** in 1960 is assumed to be in steady state considering the land use pattern and management effects of the initialization year followed by a run-in period of 15 years, till the analysis of the results start. That is inline with IPCC guidelines calling for a run-in period of 5-20 years. However, the results are heavily influenced by it as \@ref(append:initcompare) suggested. Not only the level of the SOC gap is shifted, also the declining trend is much smaller or even not visible at all anymore. For the artificial assumption of starting with potential natural vegetation in 1961 the SOC gap is fluctuating around 10 GtC for the period 1975--2010. For the more moderate assumption, that half of the transition towards the steady-state under land use has already appeared in 1960, the SOC gap is around 23 GtC in 2010. The relative declining trend for the SOC gap between 1975 and 2010 (-12%) is however almost the same as for our historic estimation (-14%). According to LUH [@hurtt_harmonization_2020] cropland area has increased over 62% since 1900 leaving our choice of initialization values too low, since reaching a new steady-state can take decades or even centuries. Having no yield or management data available before our initialization year leaves us with imperfect initialization knowledge, where our initialization choice is probably closer to real world values than starting from a natural vegetated steady-state.

Following the IPCC guidelines this study has limited its focus to the **first 30 cm of the soil profile**. In this regard several aspects are over simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Consequently, our SOC loss estimates are too high.
Secondly changes to the subsoil due to tillage are neglected, which might be important to noticed, hence studies (see Don on tillage) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. It has been argued that for intensively tilled soils, subsoil SOC is increasing due to the import of carbon rich topsoil to deeper soil layers. Following this arguments SOC stocks in croplands might even be underestimated.

Carbon displacement via **leaching and erosion** is neglected in this study. As pointed out by several studies (cite), the final fate of leached or eroded carbon is uncertain. Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focusing especially on SOC emissions, displaced but not emitted SOC can be treated as SOC stayed on croplands.

>>>>

## Agricultural management matters

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt considerably compared to constant management assumptions. This result remains robust to variations of central model parameters and initialization stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to net reduction of the SOC debt.

The evaluation of SOC stock gains and losses, however, is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 the baseline and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see [@pugh_simulated_2015]), global croplands have acted as a carbon sink according to our study. Annual C sequesteration rates of 0.2 per mille are well below the promoted 4 per mille ([@minasny_soil_2017]), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

## SOC debt and SOC drivers in literature

According to [@sanderman_soil_2017] SOC debt since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. Our estimate of 22 GtC in 2010 for cropland debt is higher as [@sanderman_soil_2017] estimations. However, there are large uncertainties in modeling SOC at the global scale, and the [@sanderman_soil_2017] pointed out that their results might be conservatively low compared to experimental results, suggesting that our results are in an acceptable range.

Furthermore, [@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to time-variant agricultural management. [@pugh_simulated_2015] considered management effects like tillage and residue returning in a static way, but neither changes over time nor alignment to observed historical data like yields-levels or no-tillage areas were taken into account. The study moreover concludes that crop productivity gains (increasing yield levels by 18%) do not lead to a substantial decline in SOC debt (less than 1% change).

Our study for the first time uses a dynamic management dataset as driver for SOC dynamics. We show that the moderate global cropland expansion of around 11% between 1974 and 2010 and the resulting depletion of SOC stocks in converted cropland has been outweighed by improvements in agricultural productivity and practices. This is in contrast to [@pugh_simulated_2015] findings of only small effects due to improved practices.

Moreover, our sensitivity analysis indicates that (1) higher returned residue biomass driver by yield increases play the most important role, followed by (2) greater manure applications due to an increasing livestock sector as well as improved animal waste management systems, and (3) the adoption of no tillage practices. Unfortunately, disentangling drivers of higher residue and manure inputs remains unsolved within this study.

## Limitations and uncertainties

Modeling management effects at the global scale, however, comes with parametric and structural uncertainties. High SOC gains in the first 30 cm of the soil compared to natural vegetation in e.g. the temperate croplands of Central Europe and UK, indicate suspicious high estimations of SOC inputs. As pointed out by [@keel_large_2017] and [@smith_how_2020], carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. [@keel_large_2017] question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries, which might not be reflected in outdated data sources. Following this line of argument, SOC results calculated in this study may overestimate actual SOC stocks due to an overestimation in residue biomass, for which the ratio compared to the harvested organ is normally reduced due to breeding. However, with an evaluation of management effects (see Sect. \@ref(sec:ipcccompare)), there is no indication for a systematic overrating of residue biomass. Rather, it is also likely that we are still missing carbon inputs e.g. from cover- and inter-cropping to the soil.

Another aspect concerns the initialization of SOC stocks in 1901, which is assumed to be in steady state considering the land-use pattern of 1901 and agricultural management data of 1965. As shown in Fig. \@ref(fig:SOCinit) the SOC debt estimation halves (from ~26 GtC to ~14 GtC) as well as the decrease in SOC debt (from ~4 GtC to ~2.5 GtC), if considering initialization SOC stocks under undisturbed natural vegetation. [@pugh_simulated_2015] pointed out the importance of accounting for the land-use history, as many CO~2~ emissions from agricultural soils are caused by historic land-use change (LUC) and the slow decline of SOC under cropland before it reaches a new equilibrium. Our results show that these legacy fluxes are compensated by agricultural management changes even under the strong assumption of the Initial-natveg scenario. However, it remains open, if our simple, depth-restricted SOC model itself can represent these fluxes correctly.

Generally, the limit to the first 30 cm of the soil profile follows the IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]) and assumes, that most of the SOC dynamic happens in the topsoil. In this regard several aspects are over simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Secondly changes to the subsoil due to tillage are neglected, which might be important to noticed, hence studies ([@powlson_limited_2014]) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. Thirdly, emissions from croplands on organic soils (like peat- and wetlands) are likely underestimate substantially.

This study excludes not only peatland degradation, it also does not account for carbon displacement via leaching and erosion. However, as pointed out by [@doetterl_erosion_2016], the final fate of leached or eroded carbon is uncertain and might even offset LUC emissions ([@wang_human-induced_2017]). Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focusing especially on the SOC debt, displaced but not emitted SOC can be treated as SOC stayed on croplands.
Moreover, the exclusive focus on croplands ignores LUC emission on other land-use types such as pastures, rangelands and forestry. Human interventions have led to large changes in SOC stocks there as well ([@sanderman_soil_2017],[@friedlingstein_global_2019]). This study does not try to be an comprehensive LUC emission analysis and acknowledges that land-use changes comes with large overall emissions.

## Modeled management effect in line with default IPCC assumptions {#sec:ipcccompare}

To validate our modeled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]) which are based on measurement data for croplands (see Table \@ref(tab:SCFtable)). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones (Fig. \@ref(append:climatemap)).

```{r captionSCFtable, echo=FALSE}
captionSCFtable <- "This table shows different estimates of stock change factors $F_{SCF}$, defined as the SOC stock of managed croplands relative to the counterfactual SOC stock under natural vegetation. We compare average values for four IPCC climate zone classifications. IPCC2006 and IPCC2019 default factors (medium input) as well as altered factors for low and high input of organic matter without any other subsystem consideration (see Table 5.5 in [@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019] are compared to 1990 and 2010 results for historic input levels of this study."
```

@@

Whereas our estimates are too low in the two tropical climate zones, temperate zone default factors are overestimated for the refinement of the IPCC guidelines ([@calvo_buendia_ipcc_2019]). Admittedly, the tropical factor differ substantially between the IPCC guidelines in 2006 ([@eggleston_hs_ipcc_2006]) and their refinement in 2019 ([@calvo_buendia_ipcc_2019]), leaving our tropical results in between.
Taking the oversimplified assumption, that tropical soils might suffer from insufficient C input rates (low) due to yield gaps , whereas temperate soils in developed regions might be highly managed and fertilized (high), the difference vanish partially. Our results show still especially in temperate dry regions -- the smallest region area-wise -- small deviations from natural SOC stocks. Considering the impact of irrigation ad fertilization on carbon-poor dryland soils, even factors above 1 (see Fig. \@ref(fig:SOCmaps)(c)) might be not surprising.

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of $F_{SCF}$ as also indicated by the time trend of the SOC debt.

## SOC stocks in line with literature

The world's SOC stock and its changes are highly uncertain, which is seen in the wide range of global SOC stock estimates for the first 30 cm of the soil profile ([@batjes_harmonized_2016], [@hengl_soilgrids250m_2017], [@fao_global_2018], [@schaphoff_lpjml4_2018-1]) in Table \@ref(tab:SOCtable)).

```{r captionSOCtable, echo=FALSE}
captionSOCtable <- "Modeled as well as data based estimation for global SOC stock in GtC for the first 30 cm of soil aggregated over all land area. Note that SoilGrids, GSOC and WISE do not consider land-use as well as changes over time and rely on soil profile data gather over a long period of time. This makes it hard to pinpoint a specific year for these SOC estimations. In this context they will be compared to modeled data from LPJmL4 for potential natural vegetation and this study for the year 2010. LPJmL4 "
```

@@

The global estimates of SOC stock from this study are on the lower end compared to other modeled results or data driven estimates. Looking at regional results in Fig. \@ref(append:regcompare), our estimates turn out to be in good agreement for most of the world, with the largest deviations for boreal regions. To avoid that this bias influences our results, our study focuses on SOC changes on cropland. Pristine natural vegetated areas without human land management cancel out in calculation of $\Delta SOC$ and $F^\mathrm{SCF}$.

Additionally, our estimates for total SOC stocks of the world, as well as our SOC initialization, are dominated by the representation of natural vegetation, which are, however, only estimated in a basic manner. For example, we do not differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass plant types. This leaves carbon inputs and decay behavior for natural litterfall rather uncertain. The absolute values of SOC stocks and debt from land-use change have to be used with caution. As our default litter parameterization accounts for woody plant types, larger uncertainty in natural land may arise especially in less forested areas.

We conducted a sensitivity analysis (Fig. \@ref(append:senslitterpnv) based on various plant parameterizations from the Century model (see Sect. \@ref(sec:scen_litterpnv)). This shows that the general trend of decreasing SOC debt of ~4 GtC is not altered under various estimates for natural SOC stocks.
